```{r echo=FALSE}
library(lattice)
opts_chunk$set(dpi = 144)
#opts_chunk$set(fig.width = 10)
#lattice.options(default.args = list(as.table = TRUE))
#lattice.options(default.theme = standard.theme)
```

```{r echo=FALSE}
data(iris) # load data from the datasets package
set.seed(13)
iris$X <- iris$Sepal.Length[seq_len(nrow(iris) / 3)] + rep(rnorm(nrow(iris) / 3), 3)
```

```{r splom, dev="png", echo=FALSE}
#splom(~ iris[-5])
```

```{r reference, dev="png", echo=FALSE}
#plot(xyplot(Sepal.Length ~ X, data = iris, groups = Species, ylim = range(iris[-5])), split = c(1, 1, 2, 2), more = TRUE)
#plot(xyplot(Sepal.Width ~ X, data = iris, groups = Species, ylim = range(iris[-5])), split = c(2, 1, 2, 2), more = TRUE)
#plot(xyplot(Petal.Length ~ X, data = iris, groups = Species, ylim = range(iris[-5])), split = c(1, 2, 2, 2), more = TRUE)
#plot(xyplot(Petal.Width ~ X, data = iris, groups = Species, ylim = range(iris[-5])), split = c(2, 2, 2, 2), more = TRUE)
```

## Performing sequential reshapes

The prerequisite of our goal (the trellis plot of our extended version of the *Iris* data) is to reshape the data into the completely long format either taking the *F.D* or the *D.F* sequence of two reshapes. The next two code blocks represent the *F.D* sequence.  The first reshape is according to *floral part* and the result is assigned to `iris.F`.

```{r}
iris.F <-
    reshape(iris, direction = "long",
            varying = list(c("Sepal.Length", "Petal.Length"), c("Sepal.Width", "Petal.Width")),
            v.names = c("Length", "Width"),
            timevar = "Floral.Part", times = c("Sepal", "Petal"),
            idvar = "id1")
```

Note how the `varying` argument is a list of equal-length vectors whose length is the number of levels (in this case two) of the factor (*floral part*) according to which we perform the reshape.  That the length of the list is also two is merely a coincidence.  In general the list may be both shorter or longer than each of its own vector components.  The concluding section will discuss how that is related to the crossed factors of the experiments.

The second reshape in the *F.D* sequence is according to *direction* yielding data frame assigned to the name `iris.F.D`.

```{r}
iris.F.D <-
    reshape(iris.F, direction = "long",
            varying = c("Length", "Width"), v.names = "Size",
            timevar = "Direction", times = c("Length", "Width"),
            drop = "id1")
```

As we see, `iris.F.D` structures data in the desired long-format (compare with the most recent table):
```{r}
head(iris.F.D, n = 2)
tail(iris.F.D, n = 2)
```

The next code block performs the reverse sequential reshape *D.F* i.e. first according to *direction* and then according to *floral part*.

```{r}
iris.D <-
    reshape(iris, direction = "long",
            varying = list(c("Sepal.Length", "Sepal.Width"), c("Petal.Length", "Petal.Width")),
            v.names = c("Sepal", "Petal"),
            timevar = "Direction", times = c("Length", "Width"),
            idvar = "id1")
iris.D.F <-
    reshape(iris.D, direction = "long",
            varying = c("Sepal", "Petal"), v.names = "Size",
            timevar = "Floral.Part", times = c("Sepal", "Petal"),
            drop = "id1")
```

The "head" and "tail" of the new data frame `iris.D.F` are the same as those of `iris.D.F` (up to the order of the `Floral.Part` and `Direction` components).

```{r}
head(iris.D.F, n = 2)
iris.D.F[201, ]
tail(iris.D.F, n = 2)
```

But the middle portion of the two data frames differ due to the different permutation of crossed factors.  That difference, however, has no statistical meaning similarly to the order of columns.  Both types of difference are implementation details that well designed software tools should be able to ignore.
```{r}
iris.F.D[201, ]
iris.D.F[201, ]
```

## The goal attained

Having sequentially shaped the *Iris* data---supplemented with our covariate $X$---into two long-format data frames, we are ready to produce the trellis display.  Starting with `iris.F.D` (resulting from the "forward" sequence)
```{r correct.F, dev="png"}
xyplot(Size ~ X | Direction * Floral.Part, data = iris.F.D, groups = Species, auto.key = list(columns = 3))
```

Using `iris.D.F` ("reverse" sequence) yields
```{r correct.D, dev="png"}
xyplot(Size ~ X | Direction * Floral.Part, data = iris.D.F, groups = Species, auto.key = list(columns = 3))
```

The two plots are identical since the differences between `iris.F.D` and `iris.D.F`---or equivalently between the corresponding sequences of reshapes---are immaterial from the viewpoint of `xyplot`, as discussed earlier.

## How to call `reshape` and why so?

The crucial but poorly documented feature of `reshape` is how "time"-varying variables must be specified for the `varying` argument in factorial experiments like the *Iris* study.  Suppose there are $k$ factors  $a_1,...,a_k$, whose levels count $p_{a_1},...,p_{a_k}$ (as we saw these factors are equivalent to index sets $\mathcal{V}_{a_1},...,\mathcal{V}_{a_k}$).  With the notation we introduced for longitudinal setups $\prod_{i=1}^k p_{a_1} = p\le n$ because these $k$ factors are fully crossed.  The remaining $q\equiv n - p$ variables, if any, are not crossed with the previous $k$ factors and consequently play a passive role in any of the reshapes.  In our extended *Iris* example $p_\mathrm{sp} = 3, \; p_\mathrm{fp} = 2, \; p_\mathrm{dir} = 2$ and $q=1$ because there is a single uncrossed variable: covariate $X$.

To illustrate the sequential use of `reshape` we will

1. reshape our extended `iris` into a completely wide format `iris.wide`
1. then we will reshape `iris.wide` according to *species* to obtain `iris.2`
1. we will check if `iris.2` is any different from `iris` (it should not be apart from implementation details)
1. discuss the semantics behind the syntax of calling `reshape`

### Reshape `iris` to a completely wide format

It will be convenient to assign the value of the `varying` argument to reshape to a name, say `varying.v`.
```{r}
(varying.v <-
    lapply(c(paste0(".Sepal", c(".Length", ".Width")),
             paste0(".Petal", c(".Length", ".Width"))),
           function(x) paste0(c("Setosa", "Versicolor", "Virginica"), x)))
```
That `varying.v` is indeed the correct argument will be shown and discussed shortly.

Additionally, we create an `id` variable, whose role is to identify all $m$ observations in the sequence $1,...,m$ repeated for all $3$ levels of *species*.
```{r}
iris$id <- rep(seq_len(nrow(iris) / 3), 3)
```

Now, the syntax of the desired reshape is this:
```{r}
iris.wide <-
    reshape(iris, direction = "wide",
            varying = varying.v,
            v.names = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
            timevar = "Species", times = c("setosa", "versicolor", "virginica"),
            idvar = "id")
```

The structure of the new data format looks as expected
```{r}
str(iris.wide)
```

### Reverse: from `iris.wide` to `iris.2`

Next, the reverse reshape operation is carried out.  The syntax is conveniently the same as the forward equivalent except for the `direction` argument, which has now the value `"long"`.
```{r}
iris.2 <-
    reshape(iris.wide, direction = "long",
            varying = varying.v,
            v.names = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
            timevar = "Species", times = c("setosa", "versicolor", "virginica"),
            idvar = "id")
```

### Checking consistency between `iris` and `iris.2`

In fact, `iris` and `iris.2` differ in their attributes, the order of their components, and in the fact that `Species` is a factor in `iris` but a character vector in `iris.2`:
```{r}
str(iris)
str(iris.2)
```

But do they differ in any meaningful way?  Reordering `iris.2` and converting its `Species` component into factor shows
```{r}
iris.2$Species <- as.factor(iris.2$Species)
all.equal(iris, iris.2[names(iris)])
```
Thus, the answer is: `iris.2` does not have any meaningful differences from `iris`, demonstrating the reversibility of reshape operations with the `reshape` function.  This means that we can perform a sequence of three reshapes from `iris.wide` and reach essentially the same `iris.F.D` or `iris.D.F` as we did with two reshapes from `iris`.

### The semantics and syntax of `reshape` and its `varying` argument

When our data are in a completely wide format, a sequence of $k$ reshapes must be used to reach the (completely) long format.  Clearly, the sequence of reshapes follows a chosen sequence of factors $a_1,...,a_k$.

First we want to reshape according to factor $a_1$ (equivalently index set $\mathcal{V}_{a_1}$.  With what arguments should we call `reshape`?  In particular what should be the value of the `varying` argument?  The above examples show that `varying` must be a list of character vectors each of length $p_{a_1}$ (see e.g. `varying.v` in the previous section).

This is because all $p_{a_1}$ levels of factor $a_1$ must appear in each vector.  The number of such vectors---the length of the list itself---is $p / p_{a_1} = \prod_{i=2}^k p_{a_i}$ because we need to consider all combinations of the *remaining* $k-1$ factors.  In case of `varying.v` $a_1 = \mathrm{sp}$ and $p_{a_1}=3$ corresponding to $\{\mathrm{setosa},\mathrm{versicolor},\mathrm{virginica}\}$, this is why each component of the list `varying.v` is a vector of length $3$.  On the other hand, $a_2 = \mathrm{fp}$ and $a_3 = \mathrm{dir}$ so $p / p_{a_1} = p_{a_2} p_{a_3} = 4$ because each of *floral part* and *direction* has $2$ levels, and this is why the length of `varying.v` itself is $4$.

In the second reshape $a_1$ plays no more role so it is omitted from `varying`.  Now the list has $p / (p_{a_1} p_{a_2}) = \prod_{i=3}^k p_{a_i}$ vector components and each vector is $p_{a_2}$ long.  The sequence continues with $i=3,...,k$.

If we follow this rule then the last, $k$-th, reshape is special in the sense that no more factors remain and hence we arrive at the nonsense product $\prod_{i=k+1}^k p_{a_i}$ (nonsense since it has zero terms).  Then `varying` is a list with a single component, a $p_{a_k}$-length vector.  Such a trivial list may be `unlist`ed: replaced by its only component (the $p_{a_k}$-length vector) without any loss of information.  Notice how we could write `varying = c("Length", "Width")` instead of (the also correct) `varying = list("Length", "Width")` when we reshaped `iris.F` into `iris.F.D`!

In fact, this "final" reshape is probably by far the most frequent way in which `reshape` is called since it was designed for longitudinal experiments where with only $k=1$ "factor" present: time.

## The lessons learned

The multiplicity of data representations is not only computationally convenient but also facilitates the clarification and expression of certain semantic relationships between variables.  Unlike `stack` the `reshape` function allows representation of various experimental design by partitioning variables into related sets and indexing those separately.  We only briefly mentioned how longitudinal experiments can be modeled with `reshape` and how this can be generalized to cases where instead of time points genes or other objects serve as indices to a set of related variables. `reshape`'s documentation provides details and examples on---mostly non-generalized---longitudinal applications.

In contrast, we examined more deeply how `reshape` is also capable of dealing with the equally important but more complex case of factorial experiments, where $k$ number of combined index sets correspond to $k$ crossed factors.  In the *Iris* data set the factors are *species*, *floral part*, and measurement *direction* so $k=3$.  Moreover, we supplemented `iris` with a covariate $X$ with the intention of $X$ having a relation to the original variables akin to time-invariant variables relate time-varying ones in longitudinal studies in a sense that data reshaping is only governed by the latter.  This demonstrated how flexibly `reshape` can be adapted to various setups.

We also saw that as many reshapes can be performed as many factors are present.  Therefore various *sequences* of reshapes may transform a completely wide-format data frame to a completely long-format one.  Thus, intermediate formats exist whenever we have $k>2$ factors.  Which sequence of reshapes we chose is immaterial, or at least should be, from the viewpoint of well-designed software such as the trellis display plotting `xyplot`.
